---
title: "lab_08"
author: "derek willis"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key

```{r echo=FALSE}
library(tidycensus)
census_api_key("847bfc233d5034b7e8a7fab09e715b8cc1872dc9")
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2019 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2019 5-year ACS and save them to a dataframe

```{r}
variables <- load_variables(year = 2019, dataset = "acs5")
```

## Load PPP data

You'll also need to load an RDS file of PPP loan applications that were removed from the original data by lenders in July.

**Task** read the RDS file and save it to a dataframe

```{r}
ppp_loans_removed <- read_rds("data/lab_08.rds")
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2019 5-year estimates? You will need to find the variable for median age first in your list of ACS variables.
**A1**
Baltimore City

```{r}
md_median_age <- get_acs(geography = "county", variables = "B01002_001", year = 2019, state = "MD") %>% arrange(estimate)
md_median_age
```

**Q2** Which Maryland counties have a median age of at least 50? Write code to isolate just those jurisdictions.
**A2**
Talbot and Worcester Counties

```{r}
md_median_age %>% 
  filter(estimate >= 50)
```

**Q3** Which Maryland counties have the highest percentage of 5-17 year-olds that speak Spanish at home? You'll need one dataframe with county-level data on the number of 5-17 year-olds that speak Spanish at home and a second dataframe with the total population of 5-17 year-olds by county. For each dataframe, rename the `estimate` column to a meaningful name you've given the ACS variable, remove the `variable` columns in each using select and then join them based on GEOID and NAME. Then you can add a column to calculate the percentage of the total 5-17 population that speaks Spanish at home.
**A3**
Prince George's and Montgomery County have the highest percentage of spanish speaking 5-17 year-olds.

```{r}
spanish_5_17 <- get_acs(geography = "county", variables = "B16007_004", year = 2019, state = "MD") %>% rename(spanish_5_to_17 = estimate) %>% select(-variable)
all_5_17 <- get_acs(geography = "county", variables = "B09001_009", year = 2019, state = "MD") %>% rename(age_5_to_17 = estimate) %>% select(-variable)

all_5_17 %>% 
  inner_join(spanish_5_17, by = c("GEOID", "NAME")) %>% 
  summarise(NAME,
            percent_spanish = spanish_5_to_17/age_5_to_17 * 100
            ) %>% 
  arrange(desc(percent_spanish))
```

**Q4** Given the counties with the largest percentage of 5-17 year-olds that speak Spanish at home, what stands out to you about the data? How would you describe the top 5 counties? How would you describe Garrett County's numbers?
**A4**
The biggest thing that stands out is that PG County is listed as having over a 100% spanish speaking population, which is not possible. Even when taking into account the margin of ever, it seems like this percentage is being overestimated. Garrett County's percentage of spanish speaking children is by far the smallest and less than half of Calvert County which is the next lowest.

**Q5** Which 5-digit zip code had the most removed PPP applications, and what is the percentage of Black population in that zip code according to the ACS? To answer this question, you will need to create separate dataframes with ACS data on total population by zip code and Black population by zip code. In each dataframe, rename the `estimate` column and remove the `variable` column. You'll need to combine the two dataframes in order to calculate the percentage, and then you can combine it with the PPP data. Remember: the PPP data may not have only 5-digit zip codes!
**A5**
Zip code 60620 in the South Side of Chicago area had the most removed PPP applications. The population in this zip code is almost 96% Black.

```{r}
top_removed_zips <- ppp_loans_removed %>% 
  mutate(zip5 = substr(zip, 1L,5L)) %>% 
  group_by(zip5) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
  #zip code 60620

black_60620 <- get_acs(geography="zcta", variables = "C02003_004", zcta = 60620, state = "IL") %>% 
  select(-variable) %>% rename(black = estimate)
total_60620 <- get_acs(geography = "zcta", variables = "C02003_001", zcta = 60620, state = "IL") %>% 
  select(-variable) %>% rename(total = estimate)

total_60620 %>% 
  inner_join(black_60620, by = c("GEOID", "NAME")) %>% 
  mutate(black_percent = black/total*100)

```

**Q6** How would you describe to an editor the top 10 zip codes with the most removed applications in terms of Black population? Where are they located, and what do they have in common? Are there any zip codes that seem different, and if so why?
**A6**
The top 10 zip codes all have large Black populations. The majority are concentrated in Chicago, but there is also one zip code in Atlanta as well.

**Q7** What other columns from the PPP data could be useful in reporting on what happened to these removed loans?
**A7**
Other useful columns could be the lender, the amount of the loan, and the approval date to get a sense of when these loans were applied for in the process, who gave them out and ultimately removed them, and how much money ultimately not given out.
