---
title: "lab_01"
author: "sean mussenden"
date: "8/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the slice of PPP loans for West Virginia.

**Task** Run the codeblock below to load the data.

```{r}
wv_ppp_loans <- read_rds("lab_01.rds")
```

## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.
**Q1** How many rows and columns does this dataframe have?
**A1** 33,336 Rows and 58 Columns

```{r}
glimpse(wv_ppp_loans)
```

## Answer questions

Use the pre-lab and your reference notebook as examples.

**Q2** Which West Virginia county got the most loans? How many?
**A2** Kanawha County with 4,273 loans.

```{r}
wv_ppp_loans %>%
  group_by(project_county_name) %>%
  summarise(
    count_loans = n()
  )%>%
arrange(desc(count_loans))

```
**Q3** Which West Virginia county got the fewest loans? How many?
**A3** Wirt County with 36 loans (Palm Beach and Allegheny are not actual counties in West Virginia).

```{r}
wv_ppp_loans %>%
  group_by(project_county_name) %>%
  summarise(
    count_loans = n()
  )%>%
  arrange(count_loans)

```
**Q4** What's your best educated guess to explain why the county identified in question 2 had a lot of loans and why the county identified in question 3 had only a few loans? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.  
**A4** Kanawha is the most populous county in West Virginia and also home of the state capital while Wirt county is the least populous in the state.


**Q5** What West Virginia business got the state's largest PPP loan? How much was it for? What does the business do? How many jobs did the loan retain, according to the data?
**A5** Greenbriar Hotel Corporation received the largest PPP loan in the amount of $8,912,202. They operate a luxury resort in the Alleghany Mountains of West Virginia. According to the data this allowed them to retain 500 jobs.

```{r}
wv_ppp_loans %>%
  group_by(name, business_type, jobs_retained, amount)%>%
  summarise(
    count_loans = n()
  )%>%
  arrange(desc(amount))

```

**Q6** What West Virginia politician (with his family) owns the company identified in question 5? What's your source for that information? Hint: this will require web research, you won't be able to answer this purely from the data set.  You do not need to write any code for this question, so no codeblock is provided.  
**A6** The current Governor of West Virginia, Jim Justice, and his family company Justice Family Group LLC own the resort. I found this information from [this article] (https://www.bizjournals.com/washington/stories/2009/05/04/daily84.htmlhttps://www.bizjournals.com/washington/stories/2009/05/04/daily84.html)

**Q7**. Produce a summary table that has the following information for West Virginia as a whole (the whole state, not by county):
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings.

**A7** The smallest loan size being only $165 seems strange. If the intention of the program is to preserve jobs, that seems like an amount of money unlikely to help retain staff. The distribution of loan amounts also seems skewed. The average loan amount was $78,017, but half of West Virginia businesses were given loans of $20,832 or less. This, combined with the reported maximum loan size of almost 9 million dollars suggests that the largest loan amounts were concentrated among a relatively small number of businesses. What industries those businesses participate in and how they used their loans could be a topic worth exploring.
```{r}
wv_summary <- wv_ppp_loans %>%
  summarise(
    count_loans = n(),
    total_loans_amount = sum(amount),
    mean_loan_amount = mean(amount),
    median_loan_amount = median(amount),
    min_loan_amount = min(amount),
    max_loan_amount = max(amount)
  ) %>%
  arrange(desc(max_loan_amount))
```
