---
title: "lab_04"
author: "derek willis"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
```

## Load Data

Let's keep working with Maryland loan applications contained in the PPP data.

**Task** Read in the lab_04.rds file and assign it to the variable `maryland_ppp`.

```{r}
maryland_ppp <- read_rds("lab_04.rds")
```

To answer the questions below, you'll need to join both the NAICS code data and the ZCTA demographic data from the pre-lab. There's a particular set of steps you'll need to perform to make that work properly, so take your time and check your work.

First, let's load the NAICS code CSV file we have in the data folder:

**Task** Read in the naics_codes.csv file and assign it to the variable `naics_codes`.

```{r}
naics_codes <- read_csv("data/naics_codes.csv")
```

Next join the two and then create a 5-character `zip5` column that we can use to join to the ZCTA data:

**Task** Write code to join the maryland_ppp and naics_codes data, assigning the result to a new variable. Then add a new `zip5` column.

```{r}
maryland_ppp_with_naics <- maryland_ppp %>% 
  left_join(naics_codes) %>% 
  mutate(zip5 = str_sub(zip, start = 1L, end = 5L))
```

Let's add the Zip Code Tabulation Area dataset. You'll need to make the `ZCTA5N` column a character column, not a numeric one.

**Task** Read in the Zip Code Tabulation Area dataset in the data folder and change the datatype of the `ZCTA5N` column to character. IMPORTANT: you can see what the column names represent by checking the data documentation, which can be found here: https://geodata.md.gov/imap/rest/services/Demographics/MD_CensusData/FeatureServer/1. You will need to refer to this to answer the questions below.

```{r}
maryland_zcta <- read_csv("data/maryland_zcta.csv") %>% 
  mutate(across(ZCTA5N, as.character))
```

Now we can join the PPP data to the ZCTA data:

**Task** Join the Maryland PPP data (with NAICS already joined) with the ZCTA data, specifying the join column in both dataframes.

```{r}
maryland_ppp_with_naics_zcta <- maryland_ppp_with_naics %>% 
  left_join(maryland_zcta, by=c("zip5" = "ZCTA5N"))
```

## Answer questions

**Q1** What are the top 5 industries (using the NAICS title) with the most PPP applications in Maryland?
**A1** The top 5 industries are Taxi Service, Beauty Salons, Full-Service Restaurants, Real Estate Agents/Brokers, and All Other Personal Services.

```{r}
maryland_ppp_with_naics_zcta %>% 
  group_by(title) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
```

**Q2** What are the top industries (using the NAICS title) with the most PPP applications in Maryland where the loan amount is more than $150,000?
**A2** For loan amounts greater than $150,000 the top industries are Full-Service Restaurants, Physicians, Plumbing/Heating/Air Conditioning Contractors, Dentists, and Commercial/Industrial Construction.

```{r}
maryland_ppp_with_naics_zcta %>% 
  filter(amount > 150000) %>% 
  group_by(title) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
```

**Q3** Which Maryland zip code with at least 20 percent non-Hispanic Asian population had the largest sum of PPP loans? Also show the number (count) of loans.
**A3** Zip code 20850 had the largest sum of PPP loans with 3,185 loans given for a total amount of \$474,525,058.

```{r}
maryland_ppp_with_naics_zcta %>% 
  filter(PNHA >= 20) %>% 
  group_by(zip5) %>% 
  summarise(
    count = n(),
    sum = sum(amount)
  ) %>% 
  arrange(desc(sum))

```

**Q4** Which Maryland zip code had the largest sum of loans where the reported race of the applicant was "Asian", and what is the percentage of non-Hispanic Asian population in that zip code? Also show the number (count) of loans.
**A4** Zip code 20850 had the largest number of applications where the applicant's race was listed as Asian. The zip code's population is 23.5% Asian. 240 loan applications in the zip code listed Asian as the applicant's race.

```{r}
maryland_ppp_with_naics_zcta %>% 
  filter(race == "Asian") %>% 
  group_by(zip5, PNHA) %>% 
  summarise(
    count = n()
    ) %>% 
  arrange(desc(count))
```

**Q5** What might explain why 20850 has a significant Asian population (23.5%) and only 240 of the 3,186 loans (7.5%) are to Asian applicants?
**A5** It's possible that even though there is a significant Asian population in the zip code, a significant number of those Asian residents may not be business owners. The high number of applicants who left the race question unanswered could also be an explanation.

``` {r}
maryland_ppp_with_naics_zcta %>% 
  filter(zip5 == 20850) %>% 
  group_by(race) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
```


**Q6** Check to see if any Maryland zip codes that have loans where the PPP data lists the applicant as being Asian BUT are located in a zip code that does not appear to have Asian population according to the ZCTA data. How many loans do they have and what is the total amount? Take a look at these zip codes - what could explain this?
**A6** There are 13 loans for a total amount of \$7,253,408 where the applicant is listed as Asian, but the NAICS data shows no Asian population. This could be the result of the business owner living in a separate zip code than the location of their business.

```{r}
zips_no_asian <- maryland_ppp_with_naics_zcta %>% 
  filter(race == "Asian" & PNHA == 0) %>% 
  group_by(zip5) %>% 
  summarise(
    num_loans = n(),
    total_amount = sum(amount),
    ) %>% 
  arrange(desc(num_loans)) 

zips_no_asian %>% 
  summarise(
    num_loans_all_zips_no_asian = sum(num_loans),
    total_amount_all_zips_no_asian = sum(total_amount)
    )
```

**Q7** How many industries have exactly one loan in the data (hint: you can filter _after_ using `summarise` on any totals you create). From a newsworthiness standpoint, which of these is most worthy of further exploration, and why?
**A7** 44 industries had only 1 PPP loan application. "Executive and Legislative Offices, Combined" would be the most newsworthy example in my opinion as it would represent a public entity taking money designed to support private businesses.

```{r}
maryland_ppp_with_naics_zcta %>%
  group_by(title) %>% 
  summarise(count = n()) %>% 
  filter(count == 1)
```
